Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
Updates from Overleaf

\documentclass[twoside]{article}

\usepackage{silence}
% Disable all warnings issued by latex starting with "You have..."
% This is a hack, there is no clean solution for that unless we install locally the package
\WarningFilter{latex}{You have requested package}

\pdfsuppresswarningpagegroup=1 %removes the warning
% This is a hack, the other way to remove is to convert all pdf into postscript or eps.

\usepackage{aistats/aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%\usepackage[accepted]{aistats2022}

\setlength{\footskip}{3.31pt}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{figs/}}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

\hypersetup{ % SLJ: my standard paper setup...
	pdftitle={MAP convergence},
	pdfkeywords={},
	pdfborder=0 0 0,
	pdfpagemode=UseNone,
	colorlinks=true,
	linkcolor=blue, %mydarkblue,
	citecolor=blue, %mydarkblue,
	filecolor=blue, %mydarkblue,
	urlcolor=blue, %mydarkblue,
	pdfview=FitH,
	pdfauthor={Anonymous},
	% draft,
	final,
}

@@

\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\nat}{\theta}
\newcommand{\m}{\mu}
\newcommand{\meanp}{\m}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

@@

@@

\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}

@@

\begin{abstract}
We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood,
yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

\section{Motivation}
\label{sec:motivation}

{\bf Exponential families} are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is a convex function and $\nat\in|Theta$ is a vector called \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
or variants of principal component analysis \citep{collins2001generalization}, among other applications.

{\bf Estimators.}
In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is known as moment matching.
Given a conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
%\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation. WDYT?}
% I think it's okay to have it only later, and refernece it here.

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
}
has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
Hence the natural question:

%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.

%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}

\begin{equation}
\boxed{
\begin{aligned}
	\textit{Can we upper}&\textit{ bound  the expected value of} \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} })?
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We know asymptotic answers.
We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}).
But in general, this problem is difficult. In this paper, we show some partial solutions, and give some idea on how to solve it.
%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.

%\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
%RLP right, not sure about how to fix this. It is hard to decide on the right amount of space for optimization versus statistics.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
A convergence rate for this algorithm would translate into the upper bound we seek.
%From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g. not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
Conversely we can see this problem as a very special case to study SMD, revealing the limits of current analysis.

\textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution. Also I think its better to write this before the contribution, to motivate better why we did this work. }A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
It could also help develop accelerated optimization methods.

\paragraph{Contributions}
After formalizing the problem of upper bounding the KL of MAP estimators (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
\end{itemize}

\paragraph{Notation}
$X,\,T(X)$ and $T$ \textcolor{red}{Don't we have T = T(X), but written tuch that it's simpler? I think its better to say it that we write T instead of T(X) for simplicity, rather than saying its a random variable.} are random variables, while $x$ is a sample.
$n$ is the number of samples and $d= \dim(T)$.
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

\section{Technical background}
\label{sec:background}
In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more exhaustive introduction.

The density of an exponential family for a sample $x$ is
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
and 2) a base measure $\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The log-partition function $\logpart$ acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
\end{align}
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
Then $\logpart$ is a strictly convex function of Legendre type,
and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
\citep{barndoffnielsen2014information}.
When explicit, we write the random variable $T = T(X)$.

@@

\paragraph{Duality}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$
(see Fig.~\ref{fig:duality}), giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
\begin{figure}[t]
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}%
%
%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
% We now introduce the Bregman divergence induced by $\logpart$,
% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
The Bregman divergence induced by $\logpart$,
measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
    \label{eq:bregman_switch}
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).}
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum at $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- we have no samples from the prior -- we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP, and view the MLE as a special case.

\paragraph{Optimization}
We can rewrite~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
}
which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
\emph{MAP is an instantiation of SMD.}
Note that this analogy does not hold for the MLE %($n_0=0$)
as it does not have any initialization to start from.
\fdk{It does, you just forget the initial condition after the first step with step-size 1?}

\section{Open Problem}
\label{sec:problem}
% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
% As we will see, this is already an interesting setting.
% We wish to estimate how well the MLE or a MAP model the true distribution.

Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution
with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
We wish to quantify how well the MLE or a MAP models the true distribution.

>>>>

\documentclass[twoside]{article}
\usepackage{aistats/aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%\usepackage[accepted]{aistats2022}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{figs/}}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

\hypersetup{ % SLJ: my standard paper setup...
	pdftitle={MAP convergence},
	pdfkeywords={},
	pdfborder=0 0 0,
	pdfpagemode=UseNone,
	colorlinks=true,
	linkcolor=blue, %mydarkblue,
	citecolor=blue, %mydarkblue,
	filecolor=blue, %mydarkblue,
	urlcolor=blue, %mydarkblue,
	pdfview=FitH,
	pdfauthor={Anonymous},
}

@@

\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\nat}{\theta}
\newcommand{\m}{m}
\newcommand{\meanp}{\m}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

@@

%\hypersetup{draft}

@@

\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}

@@

\begin{abstract}
We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family.
Surprisingly, we found no general solution to this problem in the literature.
In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD),
yet it falls out of scope for modern analysis.
We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
\end{abstract}

@@

\section{Motivation}

We kind of know how to do optimization
on smooth, strongly convex functions
when there is some bound on the stochasticity,
for example in the form of a bounded variance,
say for the problem
\begin{equation} %We might want to add a feasible set for theta?
\min_\theta f(\theta)
\end{equation}
with access to stochastic gradients
$g(\theta)$ such that
\begin{equation}
	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
\end{equation}
Estimating the mean of Gaussian noise is a special case of this problem class, using
$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(- \mu, \sigma^2)$.
SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE),
and attains the $1/t$ convergence rate.

But our understanding of stochastic optimization lacks in other domains.
A generalization of the above problem, at the intersection of optimization and statistics,
is the problem of minimizing
\begin{equation}
	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
\end{equation}
When $p(x\cond\theta)$ is an exponential family distribution,
$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.

Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori. As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes. Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.

Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
\[
R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
\]
where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.

We do not have a solution. The goal of this paper
is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
in the hope that we can make progress on this question;
Can we build algorithms for stochastic estimation
when the objective function is not (close to) quadratic
and the noise is not Gaussian?

We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}. %TODO complete
However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions.
%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve

\paragraph{Contributions}
\begin{itemize}
	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
\end{itemize}

The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one.
Progress from statistics could bring forward research in non-euclidean optimization, or the optimization of barrier losses, for instance helping in the development of accelerated methods.
On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.

% Then;
% - description of the setting
% - statistical solutions (asymptotic, l1 or l2 error)
% - optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
% - potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
% - conclusion?

\paragraph{Notation}
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

\section{Technical background}

Exponential families are to statistics what linear models are to regression.
Many classical model are exponential families, for instance the Gaussian or Bernoulli distributions. Their convexity or modeling flexibility make them one of the most widely used statistical models. In this section, we review some of their properties.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.

The exponential family for data $X\in \cX$ reads
\begin{equation}
	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
The function $\logpart$ is called the log-partition function that acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}

This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

For convenience, in this paper we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}). Also, when explicit, we write the random variable $T = T(X)$.

@@

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.

We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\m) = \min_{\nat\in\Theta}\langle \m, \nat \rangle - \logpart(\nat) \; ,
\end{align}
which matches the common notion of \textit{entropy} in information theory. In this paper, we assume that the sufficient statistic $T$ is minimal, e.g., its covariance is definite positive for all $\nat$. In such case, the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$, and moreover, $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
\[
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
\]

% The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).

\begin{figure}[t] % @remi In the figure you use mu instead of m??
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}

\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps)
We now introduce the Bregman divergence induced by $\logpart$, that measure the distance between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. Finally, it is equal to the divergence of its convex conjugate with switched arguments,
\begin{align}
	\bregman (\nat ; \nat_0)
    = \bregmanconj ( \meanp_0 ; \meanp) \; .
\end{align}

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is the number of fictive data points observed from a distribution with parameter mean $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).} 
\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

\section{Open Problem}
Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Note that $\cD$ does not have to be part of the exponential family.
Let $\nat^* := \argmin f(\nat)$.
Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
\footnote{
Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

In general, for arbitrary distributions $\cD$, it is a hard problem, especially since the KL is known for being brittle to misspecification \tocite, so let us focus on the well-specified case first $\cD = p(\cdot \cond \nat^*)$. Surely this problem has been solved, hasn't it ?
In some special cases yes, but it seems the general answer is no. 
We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.

{\bf Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
